---
title: '**R Code sample part 2**'
author: "Ivan Baksic"
date:
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 2) Downloading chromatin data and counting overlaps

In this part I downloaded chromatin data from DeepBlue epigenomic data server using DeepBlueR for seven histone modifications: H3K27ac, H3K27me3, H3K36me3, H3K4me1, H3K4me3, H3K9ac and H3K9me3.
I dowloded data for 15 ENCODE cell types and 89 Roadmap Epigenomics cell types and counted overlaps with reference genome 1Mb windows created in previous part.

Packages used
```{r}
library(dplyr)
library(tidyr)
library(DeepBlueR)
library(data.table)
library(IRanges)
library(GenomicRanges)
```

### Deep Blue data extraction and filtration

I listed Deep Blue sources.
```{r}
db_all_samples <- deepblue_list_samples()
db_all_samples$source %>% unique(.)
```
I selected only the ENCODE, Roadmap Epigenomics and DEEP (IHEC) projects
```{r}
ENCODE <- filter(db_all_samples, source == "ENCODE")
Roadmap <- filter(db_all_samples, source == "Roadmap Epigenomics")
IHEC <- filter(db_all_samples, source == "DEEP (IHEC)")
```

Since the data is not standardised, I removed redundant columns from each project data and identified biosource name column and biosample type column.  
Example for ENCODE:
```{r}
ENCODE[, .SD, .SDcols = \(x) !any(is.na(x))] %>%
  head(.)
```
For ENCODE biosource name was biosource_name and biosample type was biosample_type.  
For Roadmap biosource name was EDACC_NAME and biosample type was TYPE.  
For IHEC biosource name was biosource_name and biosample type was biomaterial_type.  

I filtered each project data by biosample type and added biosource name and biosample type columns.
```{r}
type_filter_function <- function(data, biosource_name, biosample_type) {
  type_filter <- c("primarycell", 
                   "primarytissue", 
                   "primarycelltissue", 
                   "primarycellculture", 
                   "primaryculture")
  temp <- data[[biosample_type]]
  temp2 <- data[[biosource_name]]
  result <- mutate(data, 
                   Final.type = gsub(" ", "", temp) %>%
                    tolower(.), 
                   Final.name  = temp2) %>%
    filter(., Final.type %in% type_filter) %>%
  return(result)
}

ENCODE_filtered_types <- type_filter_function(ENCODE, "biosource_name", "biosample_type")
Roadmap_filtered_types <- type_filter_function(Roadmap, "EDACC_NAME", "TYPE")
IHEC_filtered_types <- type_filter_function(IHEC, "biosource_name", "biomaterial_type")
```

I combined all project data frames into one and kept only relevant columns.
```{r}
final_samples <- bind_rows(ENCODE_filtered_types,
                           Roadmap_filtered_types,
                           IHEC_filtered_types)
# removed columns that contained any NA
final_samples <- final_samples[, .SD, .SDcols = \(x) !any(is.na(x))]
final_samples <- select(final_samples, -c("_id", "biosource_name", "user", "Final.type"))

dim(final_samples)
head(final_samples)
```
### Experiment data download

In my thesis I gathered all experiments for each epigenetic mark from DeepBlue database.  
Here, I will gather experiments for H3K9ac and H3K9me3 markers and first 50 samples:  
WARNING: This may take a minute
```{r}
epigenetic_marks = c(#"H3K27ac",
                     #"H3K27me3",
                     #"H3K36me3",
                     #"H3K4me1",
                     #"H3K4me3",
                     "H3K9ac",
                     "H3K9me3")

gather_experiments <- function(samples, marks, result_df = data.frame(), i = 1){
  
  # gather experiments for one epigenetic mark
  all_experiments <- sapply(samples$id, function (x) {
    suppressMessages(deepblue_list_experiments(genome = "hg19",
                                               epigenetic_mark = marks[i],
                                               type = "peaks",
                                               sample = x))
    }, simplify = F, USE.NAMES = TRUE)
  
  # remove all ids with no experiments and create dataframe from list
  all_experiments <- all_experiments[all_experiments != "\n"] %>%
    bind_rows(., .id = "sample_id") %>%
    mutate(., mark = marks[i])
  
  
  result <- rbind(result_df, all_experiments)
  
  if (i == length(epigenetic_marks)){
    return(result)
  } else {
    gather_experiments(samples,
                       marks,
                       result_df = result,
                       i = i + 1)
  }
}

H3K9ac_H3K9me3_experiments <- gather_experiments(final_samples[1:50], epigenetic_marks)

dim(H3K9ac_H3K9me3_experiments)
head(H3K9ac_H3K9me3_experiments)
```
I loaded the data with all experiments.
```{r}
all_experiments <- read.csv("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/all_experiments.csv")
dim(all_experiments)
head(all_experiments)
```

### Further data filtration

I removed all the cell types containing data for only one histone modification.
```{r}
filtered_sample_ids <- all_experiments %>%
  select(sample_id, mark) %>%
  unique(.) %>%
  group_by(., sample_id) %>% 
  summarise(total_count=n()) %>%
  filter(., total_count != 1) %>%
  .$sample_id

final_samples_filtered_ids <- filter(final_samples, id %in% filtered_sample_ids)

dim(final_samples_filtered_ids)
head(final_samples_filtered_ids)
```

I manually excluded certain samples and formed 11 groups based on histological relationship between certain cell types.
This resulted in the data frame containing 89 Roadmap Epigenomics cell types and 15 ENCODE cell types.
```{r}
biomarker_df <- read.csv("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/biomarker_list.csv")
dim(biomarker_df)
head(biomarker_df)
```
With this data frame ids I filtered all_experiments data frame.
```{r}
all_experiments_filtered <- filter(all_experiments, 
                                   all_experiments$sample_id %in% biomarker_df$id)

dim(all_experiments_filtered)
head(all_experiments_filtered)
```

### Deep Blue chromatin peak download

With gathered experiments I downloaded peak data from DeepBlue database for each epigenetic marker.
In this example I will only download data for first two samples of H3K9ac marker:
```{r}
# create nested list from data frame with samples as names
H3K9ac_experiments <- filter(all_experiments_filtered, mark == "H3K9ac") %>%
  split(., .$sample_id)

get_epigenom_data <- function(input, my_samples, gr = GRanges(), i = 1) {
  
  #dowload DeepBlue peak data
  data_id <- deepblue_select_experiments(input[[i]]$name)
  request_id <- deepblue_get_regions(query_id = data_id,
                                             output_format = "CHROMOSOME,START,END,NAME,SCORE,STRAND,SIGNAL_VALUE,P_VALUE,Q_VALUE,PEAK")
  request_data <- deepblue_download_request_data(request_id)
  
  #add meta columns to Grange object
  mcols(request_data)$biosample <- my_samples[my_samples$id == names(input)[i],"Final.name"]
  mcols(request_data)$group <- my_samples[my_samples$id == names(input)[i],"Final.group"]
  mcols(request_data)$mark <- unique(input[[i]]$mark)
  
  new_gr <- append(gr, request_data)
  
  if (i == length(input)) return(new_gr) else {
    get_epigenom_data(input, my_samples, gr = new_gr, i = i + 1)
  }
}

H3K9ac_ranges_1to2 <- get_epigenom_data(H3K9ac_experiments[1:2], biomarker_df)

length(H3K9ac_ranges_1to2)
head(H3K9ac_ranges_1to2)

```
### Overlap count

I counted number of overlaps between H3K9ac peak rage data and reference genome for every biosample.
```{r}
final_reference_hg19_2128 <- readRDS(url("https://github.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/raw/main/final_reference_hg19_2128.RDS"))

count_ol <- function(mark_ranges, reference_genome){
  
  # extracted unique biosample vector
  mark_ranges_biosample <- mark_ranges$biosample %>%
    unique(.)
  
  #counted overlaps for each biosample
  count <- sapply(mark_ranges_biosample, function (x) {
    temp <- mark_ranges[mark_ranges$biosample == x]
    result <- countOverlaps(reference_genome, temp)
    return(result)
    }, USE.NAMES = T, simplify = F)
}

H3K9ac_ranges_1to2_count <- count_ol(H3K9ac_ranges_1to2, final_reference_hg19_2128)

names(H3K9ac_ranges_1to2_count)
H3K9ac_ranges_1to2_count[[2]][1:100]
```
Final table with all markers and biosamples:
```{r}
all_markers_ranges_count_template_final <- read.csv("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/all_markers_ranges_count_template_final.csv")

dim(all_markers_ranges_count_template_final)
head(all_markers_ranges_count_template_final)
```



