---
title: '**R Code sample part 2**'
author: "Ivan Baksic"
date:
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 3) Downloading melanoma single based substitution mutation type (SKCA-BR) and counting overlaps

In this part I downloaded melanoma SNP data, performed filtration and counted overlaps with each reference genome window created in part 1. Finally, I merged mentioned melanoma SNP window counts with mark counts created in part 2 for every biosample.

Packages used
```{r}
library(dplyr)
library(tidyr)
library(data.table)
library(GenomicRanges)
library(ggplot2)
```

From ICGC DCC Data Portal Release 28 I downloaded melanoma mutations SKCA-BR data set and filtered it.  
Here, I filtered SKCA-BR subset as an example:
```{r}
SKCA_BR_sample <- fread("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/simple_somatic_mutation.open.SKCA-BR_sample.csv")

#filter Whole Genome Sequencing strategy and single base substitution mutation type
SKCA_BR_sample_WGS_SNP <- filter(SKCA_BR_sample,
                                 sequencing_strategy == "WGS",
                                 mutation_type == "single base substitution") %>%
  
  # create columns for GRanges
  mutate(.,
         chromosome = paste("chr", chromosome, sep = ""), 
         start = chromosome_start, 
         end = chromosome_end) %>%
  
  #select only relevant columns
  select(.,
         chromosome, 
         start, 
         end) %>%
  GRanges(.)

SKCA_BR_sample_WGS_SNP

```

Filtering whole SKCA-BR data set results in the following:
```{r}
SKCA_BR_full_WGS_SNP <- readRDS(url("https://github.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/raw/main/SKCA_BR_full_WGS_SNP_no_mcols.RDS"))
SKCA_BR_full_WGS_SNP
```
I counted overlaps between filtered dataset and reference genome windows and created count boxplot.
```{r}
final_reference_hg19_2128 <- readRDS(url("https://github.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/raw/main/final_reference_hg19_2128.RDS"))

SKCA_BR_full_WGS_SNP_count <- countOverlaps(final_reference_hg19_2128, SKCA_BR_full_WGS_SNP)

data.frame(SKCA_BR_full_WGS_SNP_count) %>%
  mutate(., label = "SKCA_BR_full_WGS_SNP") %>%
  ggplot(aes(x = label, y = SKCA_BR_full_WGS_SNP_count)) +
  geom_boxplot() +
  theme_bw() +
  theme(aspect.ratio=8/3)
```

From the boxplot I noticed that there many data outliers. In my thesis I have demonstrated that excluding outliers can significantly improve results so I removed outliers in this example.  
I added melanoma SNP data to chromatin mark data and filtered out outliers.
```{r}
all_markers_ranges_count_template_final <- read.csv("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/all_markers_ranges_count_template_final.csv")

# create nested list with biosamples as names
all_markers_ranges_count_template_final <- split(all_markers_ranges_count_template_final, all_markers_ranges_count_template_final$ind)

SKCA_BR_full_WGS_SNP_table_no_out <- lapply(all_markers_ranges_count_template_final,
                                            function (x) {
                                              # add mutation window counts
                                              mutate(x, mutation = SKCA_BR_full_WGS_SNP_count) %>%
                                                select(., -c(ind, window_id)) %>%
                                                # remove all colums with NA
                                                select_if(.,~any(!is.na(.))) %>%
                                                # filter out outlier mutation count observations
                                                filter(., mutation <= IQR(SKCA_BR_full_WGS_SNP_count)*1.5 + quantile(SKCA_BR_full_WGS_SNP_count, 0.75))
                                              })

head(SKCA_BR_full_WGS_SNP_table_no_out[[1]])
```
This reduced the number of windows to 1993.
```{r}
dim(SKCA_BR_full_WGS_SNP_table_no_out[[1]])
```


I used created list of dataframes to built machine learning models in part 4.