---
title: '**R Code sample part 1**'
author: "Ivan Baksic"
date:
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 1) Creating 1Mb reference genome windows

In this part I demonstrated reference genome filtration and division into 1Mb windows.

Packages used
```{r}
library(dplyr)
library(tidyr)
library(IRanges)
library(GenomicRanges)
library(data.table)
```


Since hg19_umap.csv (genome uniquely mappable regions downloaded from: https://bismap.hoffmanlab.org/) had 400001214 lines, I made recursive function and loaded and filtered chunks of 10M lines.
As an example, I read 24.2Mb file the same way (chunks of 100k lines):
```{r}
my_reference <- function(final_number, 
                         skip_number, 
                         my_ranges = GRanges(), 
                         my_skip = 1){
  my_chunk <- fread("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/hg19_umap_sample.csv",
                    skip = my_skip, 
                    nrows = skip_number) %>%
    .[V4 >= 0.92]
  # excluded regions with fraction of uniquely mappable 36-mers lower than 0.92
  
  my_chunk_regions <- GRanges(seqnames = my_chunk$V1, 
                              ranges = IRanges(start = my_chunk$V2, 
                                               end = my_chunk$V3)) %>%
    reduce(.)
  
  my_result <- union(my_ranges, my_chunk_regions)
  
  if (my_skip == final_number-skip_number)
    return(my_result) else
      my_reference(final_number, 
                   skip_number, 
                   my_ranges = my_result, 
                   my_skip = my_skip + skip_number)
}

Umap_k36_hg19 <- my_reference(900001, 100000)
Umap_k36_hg19
```
I additionally removed centromere and telomere regions:
```{r}
centromere_telomere_hg19 <- read.table("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/hg19_centromere_telomere", 
                                       header = F)

centromere_telomere_hg19_ranges <- GRanges(seqnames = centromere_telomere_hg19$V2,
                                           # one-based coordinate system to zero-based coordinate system transformation
                                           ranges = IRanges(start = centromere_telomere_hg19$V3 - 1,
                                                            end = centromere_telomere_hg19$V4))

filtered_reference_hg19 <- setdiff(Umap_k36_hg19, centromere_telomere_hg19_ranges)
filtered_reference_hg19
```
Algorithm that divides filtered hg19 reference to 1Mb windows:
```{r}
# function to add 160 ranges till 1Mb is surpassed
to_million_by_160 <- function(width_vector, count = 1, my_sum = 0) {
  new_sum = my_sum + sum(width_vector[count:(count+159)])
  if (new_sum >= 1000000) {
      return(list(count - 1, my_sum))
      } else {
        to_million_by_160(width_vector, count = count + 160, my_sum = new_sum)
      }
}

# function to add 1 range till 1Mb is surpassed
to_million_by_1 <- function(width_vector, count, my_sum) {
  new_sum = my_sum + width_vector[count + 1]
  if (new_sum >= 1000000)
    return(list(count, my_sum)) else
      to_million_by_1(width_vector, count = count + 1, my_sum = new_sum)
}


create_windows <- function (input_ranges, output = GRangesList()) {
  
  if (sum(width(input_ranges)) >= 1000000) {
    
    #first I add 160 ranges at the time to surpass 1Mb and then I add one at the time to surpass 1Mb
    broad <- to_million_by_160(width(input_ranges))
    narrow <- to_million_by_1(width(input_ranges), broad[[1]], broad[[2]])
    final_range <- narrow[[1]] + 1
    
    #missing bases are the ones needed to reach 1Mb
    missing <- 1000000 - narrow[[2]]
    
    if (missing != 0) {
      #I split final range that surpasses 1Mb in order to reach 1Mb and added the extra bases at the start of the next input
      final_range_start <- GRanges(seqnames = seqnames(input_ranges[final_range]),
                                   ranges = IRanges(start = start(input_ranges[final_range]),
                                                    end = start(input_ranges[final_range]) + missing - 1))
      final_range_end <- GRanges(seqnames = seqnames(input_ranges[final_range]),
                                 ranges = IRanges(start = start(input_ranges[final_range]) + missing,
                                                  end = end(input_ranges[final_range]) + missing))

      new_input_ranges <- append(final_range_end,
                                 input_ranges[-(1:final_range)])
      million_chunk <- append(input_ranges[1:narrow[[1]]],
                              final_range_start)
      
      #sometimes it can happen that the ranges add exactly to 1Mb, hence this else condition
    } else {
    new_input_ranges <- input_ranges[-(1:final_range)]
    million_chunk <- input_ranges[1:narrow[[1]]]
      }
    new_output <- append(output,
                         GRangesList(million_chunk))


    create_windows(input_ranges = new_input_ranges, output = new_output)
    } else return(output)
}

final_reference_hg19 <- create_windows(filtered_reference_hg19)
```

Reference genome divided into 4 1Mb windows and saved as GRangesList object.
```{r}
final_reference_hg19[[1]]
final_reference_hg19[[2]]
sapply(final_reference_hg19, function (x) sum(width(x))) %>%
  table(.)
```

