---
title: '**R Code sample part 4**'
author: "Ivan Baksic"
date:
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 4) Random Forest regression and result interpretation

In this part I performed Random Forest regression for a subset of biosamples as an example and displayed result plots for all biosample models. I compared Rsquared Values of different biosample models and correctly determined melanoma cell-of-origin.

Packages used
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(caret)
library(ranger)
library(forcats)
```

I loaded data frame created in part 3 that contains window counts for chromatin marks and melanoma SNP mutations for each biosample
```{r}
SKCA_BR_full_WGS_SNP_table_no_out <- readRDS(url("https://github.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/raw/main/SKCA_BR_full_WGS_SNP_table_no_out.RDS"))
head(SKCA_BR_full_WGS_SNP_table_no_out$`Adipose Nuclei`)
```

### Model building

I performed 10-fold cross validation Random Forest regression on all biosamples with chromatin mark counts as predictor variables and melanoma SNP mutation count as response variable.  
As an example I created Random Forest regression models on the first two biosamples:
```{r}
set.seed(52)
my_analysis <- function (input) {
  sapply(input, function (x) {
    control <- trainControl(method="cv", number=10,returnResamp="final", savePredictions = "final")
    result <- train(mutation ~ .,
                 data = x,
                 method = "ranger",
                 tuneLength = 10,
                 trControl = control,
                 num.trees = 100,
                 importance = "permutation",
                 metric = "Rsquared")
    return(list(max(result$results$Rsquared), result$pred, result$resample))
  }, USE.NAMES = T)
}
SKCA_BR_full_WGS_SNP_table_no_out_2model <- my_analysis(SKCA_BR_full_WGS_SNP_table_no_out[1:2])
SKCA_BR_full_WGS_SNP_table_no_out_2model
```

For each biosample there are several results.  
First row represents model Rsquared value:
```{r}
SKCA_BR_full_WGS_SNP_table_no_out_2model[1, 1] %>%
  unlist(.)
```
Second row contains data frame with observed and predicted values
```{r}
SKCA_BR_full_WGS_SNP_table_no_out_2model[[2, 1]] %>%
  head(.)
```

Third row contains RMSE and Rsquared values for each CV Fold.
```{r}
SKCA_BR_full_WGS_SNP_table_no_out_2model[[3, 1]]
```

### Biosample comparison

I loaded Random Forest regression results for all 104 biosamples.
```{r}
SKCA_BR_full_WGS_SNP_table_no_out_104model <- readRDS(url("https://github.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/raw/main/SKCA_BR_full_WGS_SNP_table_no_out_104model.RDS"))
```

I compared Rsquared values between all 104 biosamples.
```{r, fig.height = 9, fig.width = 11}
plot_biosamples_score <- function (Rsquared_result) {
  
  # extract data frame of R2 results for every biosample
  Rsquared_result_table <- Rsquared_result[1, ] %>% unlist(.) %>%
    data.frame(Rsquared = ., biosample = names(.)) %>%
    mutate(., Rsquared = Rsquared*100) %>%
    # arrange it in descending order
    arrange(., desc(Rsquared))
  
  my_samples <- read.csv("https://raw.githubusercontent.com/Baksic-Ivan/R_code_sample-Cancer_Cell-of-Origin_Determination/main/biomarker_list.csv")
  
  #add group names to result data frame from biomarker data frame based on biosample column
  Rsquared_result_gg <- my_samples %>% mutate(., biosample = Final.name) %>%
    left_join(Rsquared_result_table, ., by = "biosample")
  
  # get the position of the second best group of biosamples needed for the full horizontal line
  second_group <- filter(Rsquared_result_gg, Final.group != Final.group[1])
  second_group_pos <- nrow(Rsquared_result_gg) - nrow(second_group) + 1
  
  plot <- Rsquared_result_gg %>%
    # map biosamples to descending order of Rsquared column
    mutate(biosample = forcats::fct_reorder(biosample, dplyr::desc(Rsquared))) %>%
    
    #CREATE PLOT
    ggplot(aes(x = biosample, y = Rsquared, fill = Final.group)) +
    
    # create bar chart
    geom_col(width = 0.8) +
    
    #adjust themes and axis labels
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.3, hjust = 1, size = 9), aspect.ratio = 1/6,
          panel.background = element_rect(colour = "black"),
          legend.position = "bottom") +
    
    # add horizontal lines at median and second best biosample group
    geom_hline(aes(yintercept = median(Rsquared)), linetype = 2) +
    geom_hline(aes(yintercept = sort(Rsquared, decreasing = T)[second_group_pos]), size = 1) +
    
    # paint the bars by the biosample group
    guides(fill = guide_legend(title = "Cell type group")) +
    
    # add second y axis
    scale_y_continuous(name = "% variance explained",
                       sec.axis = sec_axis(~((. - median(Rsquared_result_table$Rsquared)))/mad(Rsquared_result_table$Rsquared), 
                                           name = "Robust ẑ-score", 
                                           breaks = c (-4, -2, 0, 2, 4, 6, 8, 10))) +
    theme(axis.title.y = element_text(size = rel(1))) +
    labs(x = "Cell type")
  
  return(plot)
}

plot_biosamples_score(SKCA_BR_full_WGS_SNP_table_no_out_104model)
```

Plot description:  
Solid horizontal line describes variance explained reported by the next-best cell type model that belong to the different cell type group. Dotted horizontal line indicates median variance explained value of all the cell type models. Right-hand (secondary) y-axis shows robust ẑ-score. Bars are colored by histological group.

From this result I can observe that biosamples Melanocyte 03 and Melanocyte 01 have considerably higher Rsquared value than the rest of the biosample models. In order to prove this I performed WMW pair test between neighbouring biosamples based on their fold Rsquared results.

```{r}
compare_biosamples <- function (input) {
  
  input_result_table <- input[1, ] %>% unlist(.) %>%
    data.frame(Rsquared = ., biosample = names(.)) %>%
    # arrange it in descending order
    arrange(., desc(Rsquared))
  
  
  biosamples_ordered <- input_result_table$biosample
  
  result <- sapply(1:(length(biosamples_ordered) - 1), function(i) {
    
    # get Rsquared for each fold for certain biosample
    t1 <- input[[3, biosamples_ordered[i]]] %>%
      arrange(., Resample) %>% .$Rsquared
    
    # get Rsquared for each fold for neighbour biosample
    t2 <- input[[3, biosamples_ordered[i + 1]]] %>%
      arrange(., Resample) %>% .$Rsquared
    
    wilcox_result <- wilcox.test(t1, t2, paired = T)
    return(list(biosamples_ordered[i], biosamples_ordered[i + 1], wilcox_result$p.value))
  })
  return(result)
}

WMW_test_biosamples <- compare_biosamples(SKCA_BR_full_WGS_SNP_table_no_out_104model)

WMW_test_biosamples[,1:5]
```
First two rows represent biosample neighbours and the third row represents WNW paired test p-value.


Next, I checked which of the pairs has statistically significant different Rsquared value.
```{r}
WMW_test_biosamples[,WMW_test_biosamples[3,] %>% unlist(.) <= 0.03]
```
The only statistically significant different Rsquared value is between biosample Melanocyte 01 Fetal Kidney.

Finally, I concluded that the cell type that fits the model with the highest prediction accuracy (Melanocyte) and significantly differs from the next-best histologically unrelated cell type (Fetal Kidney) is the correct cell-of-origin.
